---
title: "GBIF Observations For CT Bioblitz 2016 "
author: "James Mickley"
output:
  html_document:
    keep_md: yes
    theme: readable
  pdf_document:
    keep_tex: no
    latex_engine: xelatex
graphics: yes
---


```{r 'Setup', include=F}
# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@
# @@@@@ Knitr Options
# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@

# Set default knitr options: 
# Suppress warnings and messages, cache chunks, 
#  set default figure size to 6x8 at 300 dpi, and save a png and pdf
knitr::opts_chunk$set(warning = F, message = F, collapse = T, cache = T,
    fig.height = 6, fig.width = 8, dpi = 300, # 6x8" @ 300dpi:1800x2400=4.3MP
    dev = c('png', 'pdf'), dev.args = list(pdf = list(onefile = F)))

```

## Historical Species Records in Hartford

I downloaded all of the observations on GBIF in the area of our bioblitz and took a quick look at the data last night.  This may give a starting point to look for rare species, species that haven't been seen in a while, and under-sampled groups.

The original dataset is is available from GBIF [here](http://doi.org/10.15468/dl.vptagl), but has been run through [OpenRefine](http://openrefine.org/) using [these steps](openrefine-data-cleaning-steps.json) to clean it up a bit. 

Then, the distance from Two Rivers Middle School was calculated using the [haversine formula](http://andrew.hedges.name/experiments/haversine/).  

The data includes some of Yale's collections, research-grade iNaturalist observations and some of UConn's collections.  A large portion of the observations are from eBird. 

I included all complete observations within a 5 mile radius, and then tallied up the number of observations for each species.  The entire dataset of all 63,000 observations is available [here](5-mile-radius-cleaned.csv.zip).

A more useful dataset is a list of species (1,575 total), with the number of observations summarized, and the location and coordinates of the most recent observation for that species.  That dataset is available [here](species-list.csv)

Finally, I [subset the list of species](no-obs-in-50yr.csv) to only include those that hadn't been recorded in the last 50 years in this area.  That still left 928 species.  


If you'd like to play with this yourself, feel free to download the data.  You can also download this folder and run the [R code](Historic Species.Rmd).


```{r 'Data_wrangling', include=F}

# Center point: 
# 41.756886, -72.6543005

# Load required R libraries
library(cowplot) # Needed for publication-quality plots
library(dplyr) # Needed for data wrangling, load last to avoid conflicts

# Read in gbif data within 5 miles of Three Rivers
gbif <- read.csv(unz("gbif-5mi-radius.csv.zip", "gbif-5mi-radius.csv"))

# Take a look at the structure of the dataset
str(gbif)
names(gbif)



# Clean up the data a bit
gbif <- gbif %>%
     
     # Set month and day to Jan 1st, if not specified
     mutate(month = ifelse(is.na(month), 1, month), 
            day = ifelse(is.na(day), 1, day)) %>%

     # Create a date column
     mutate(date = as.Date(paste(year, month, day, sep="-"))) %>%
     
     # Rename species with infraspecific info
     mutate(species = factor(ifelse(infraspecies == "", as.character(species),
          paste(species, " ", taxonrank, ". ", infraspecies, sep="")))) %>%
     
     # Summarize the number of observations for each species
     group_by(species) %>%
     mutate(num.obs = n()) %>%
     ungroup

# Write out the cleaned dataset to CSV
write.csv(gbif, "5-mile-radius-cleaned.csv")

# Zip up the resulting large CSV file
zip("5-mile-radius-cleaned.csv.zip", "5-mile-radius-cleaned.csv")

# Delete the unzipped file
file.remove("5-mile-radius-cleaned.csv")



# Summarize by species, and sort by num.obs ascending
species <- gbif %>%
     
     # Arrange by date descending, and then keep the most recent record
     arrange(desc(date)) %>%
     distinct(species) %>%
     
     # Arrange by the rarest species first, and then the oldest observations
     arrange(num.obs, year) %>%
     
     # Select a subset of columns
     select(phylum, class, family, species, date, year, lat, lon, locality, num.obs)

# Write out the species list to a CSV file
write.csv(species, "species-list.csv")



# Find all the species that haven't been seen in 50 years
oldspp <- species %>%
     
     # Filter out dates with a year less than 1966
     filter(format(date,'%Y') < 1966) %>%
     
     # Put the historically most-observed species first
     arrange(desc(num.obs))

# Write out the species that haven't been seen lately to CSV file
write.csv(oldspp, "no-obs-in-50yr.csv")



```

## Some Quick Graphs

This graph shows the number of observations for each species by taxonomic Class.  It looks like only plants and birds are fairly well-sampled, though this misses UConn's insect collection I think.  

```{r 'Num-observ', echo = F}


# Filter out species that don't have class specified
species %>% filter(class != "") %>%
     
# Set up x and y for ggplot
ggplot(aes(x = class, y = num.obs)) +
     
     # Add points for each species
     geom_jitter(alpha = 0.5, color = "tomato") +
     
     # Add transparent boxplot
     geom_boxplot(alpha = 0) + 
     
     # Make the y-axis on a log scale, with the following tickmarks
     scale_y_log10(breaks = c(1, 2, 5, 10, 20, 50, 100, 200, 500, 1000, 2000)) + 
     
     # Add axis labels
     ylab("Number of Observations") + xlab("Class") + 
     
     # Black and white theme
     theme_bw() + 
     
     # Custom theming of the axis labels
     theme(axis.text.x = element_text(size = 16, angle = -45, hjust = 0),
          axis.text.y = element_text(size = 16),
          axis.title.x = element_text(size = 18, face = "bold"),
          axis.title.y = element_text(size = 18, face = "bold", vjust = 1.5))



```

And here's a distribution by class of the year where a species was last seen.  It looks like there are a lot of plant species in particular that have not been seen in a long time.  

```{r 'Year-dist', echo = F}

# Filter out species that don't have class specified
species %>% filter(class != "") %>%

# Make a column for the year last seen
mutate(last_yr = as.numeric(format(date,'%Y'))) %>%
     
# Set up x and y for ggplot
ggplot(aes(x = class, y = last_yr)) +
     
     # Add points for each species
     geom_jitter(alpha = 0.5, color = "tomato") +
     
     # Add transparent boxplot
     geom_boxplot(alpha = 0) + 
     
     # Add axis labels
     ylab("Last Observed (year)") + xlab("Class") + 
     
     # Black and white theme
     theme_bw() + 
     
     # Custom theming of the axis labels
     theme(axis.text.x = element_text(size = 16, angle = -45, hjust = 0), 
          axis.title.x = element_text(size = 18, face = "bold"),
          axis.text.y = element_text(size = 16),
          axis.title.y = element_text(size = 18, face = "bold", vjust = 1.5))



```


### R Session Information (for posterity)

```{r 'Session_Info', echo=F, comment=""}

# Add session information to help with reproduceability
sessionInfo()
